---
title: "S.W.E.L.T.R invertebrates"
description: |
   **S**oil **W**arming **E**xperiment in **L**owland **Tr**opical forest

author:
  - name: "Hubert Szczygieł"
    orcid_id: 0000-0002-4290-4660
    url: https://www.inaturalist.org/people/hubertszcz
    affiliation: Smithsonian Tropical Research Institute
    affiliation_url: https://stri.si.edu/
  - name: "Orpheus Butler"
    orcid_id: 0000-0001-5345-7889
    url: https://orpheusbutler.com/
    affiliation: The University of Sydney
    affiliation_url: https://www.sydney.edu.au/
  - name: "Andrew Nottingham"
#    orcid_id: 0000-0001-9421-8972
    url: https://environment.leeds.ac.uk/geography/staff/9617/dr-andrew-nottingham
    affiliation: University of Leeds
    affiliation_url: https://www.leeds.ac.uk/

output: distill::distill_article

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
url <- "https://static.inaturalist.org/photos/96719643/large.jpg"
```

<center><img src="`r url`"></center>
#### Entomobryomorpha

## Background
Climate change is exacerbating a global decline in biodiversity. Numerous observational studies link rising temperatures to declining biological richness, diversity, and abundance in terrestrial ecosystems, yet almost no studies have considered the highly diverse and functionally significant communities of soil and leaf-litter fauna associated with tropical forests. Here, we report an order-level decline in the richness and diversity of soil and leaf-litter fauna following three years of experimental whole-profile soil warming in a tropical forest. Diversity declines were greatest during the dry season, revealing that future effects of warming on tropical soil fauna could be worsened with coinciding drought events. Contrary to findings from higher latitudes, total faunal abundance increased under warming, with effects varying strongly among taxa, ranging from ~20-fold reductions in Protura to ~7-fold increases in soil-associated Annelida. These changes translated to major shifts in community composition that have likely contributed to parallel changes in soil microbial communities with negative consequences for soil carbon storage. Our study provides direct experimental evidence that warming causes diversity declines and compositional shifts for tropical forest soil and leaf-litter fauna, a result that has direct consequences for the carbon (C) cycle and highlights the sensitivity of tropical biodiversity to climate change.