---
title: "Linear-models"
description: |
  Reproducible workflow for 
output:
    distill::distill_article:
      css: assets/styles.css
      toc: true
      toc_depth: 3
---

Here we use data frames created in the [initial setup](initial.html) which contain mass-corrected count data as well as Shannon Diversity values calculated for each plot-date. Soil and leaf-litter data are treated similarly, though run seperately. First we start with soil:

<details markdown="1">
<summary><strong>Click here</strong> for setup information.</summary>

```{r setup, message=FALSE, results = 'hide'}
soil.abundance <- read.csv("soil.abundance.csv")
soil.abundance2 <- read.csv("soil.abundance2.csv")
soil.data <- read.csv("soil.data.csv")

library(ggeffects)
library(lme4)
library(ggplot2)
library(scales) 

xaxis <- c(" ", " ")
```
</details>

# Total Abundance

We started with the following model:

```{r, message=FALSE, results = 'hide'}

#moisture and plot type, log distribution
require(glmmTMB)
require(DHARMa)
require(car)

abundance.mod <-glmmTMB(integer.abundance~Avg..Soil.moisture*Plot.Type+(1|Plot.pair),
                        data=soil.data,
                        ziformula=~0,
                        family=nbinom2(link="log"))	 #this could be poisson or various other distributions
abundance.resids<-simulateResiduals(abundance.mod)
abundance.resid.plot<-plot(abundance.resids)
```

But ended up using this one:
```{r, message=FALSE, results = 'hide'}
abundance.mod <- glm(integer.abundance ~ Avg..Soil.moisture * Plot.Type, family = poisson, data=soil.data)
```

We visualized the difference in total abundance between warmed and control plots using box plots with violins to show distribution of data:

```{r, echo=FALSE, message = FALSE, warning = FALSE}
ggplot(soil.data, aes(x = Plot.Type, y = abundance, color = Plot.Type, group = Plot.Type)) + 
  scale_fill_manual(values = c("royalblue1", "tomato1")) + scale_color_manual(values = c("royalblue1", "tomato1")) +
  geom_violin(trim=FALSE)+  theme_minimal()+ labs(y= "Total abundance", x = "")+ geom_boxplot(size=0.1) +
  theme(plot.title = element_text(hjust = 0.5)) + stat_summary(fun=mean, geom="point") + stat_summary(fun.data = mean_se, geom = "errorbar", aes())+
  theme(legend.position="none",legend.text=element_text(),legend.direction='horizontal',plot.title = element_text(hjust=0.5),text=element_text(family="Arial"))+
  scale_x_discrete(labels= xaxis) +
  theme(text=element_text(family="Arial")) + 
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "transparent",colour = NA),
    plot.background = element_rect(fill = "transparent",colour = NA))+
  scale_y_continuous(labels=comma)+
    theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))

```

We visualized the interaction of the warming and moisture effects using an effects plot:

```{r, echo=FALSE, message = FALSE, warning = FALSE}
abundance.effects <- ggpredict(abundance.mod, terms = c("Avg..Soil.moisture", "Plot.Type"))

ggplot(abundance.effects, aes(x = x, y = predicted, colour = group)) + 
  scale_fill_manual(values = c("tomato1","royalblue1")) + scale_color_manual(values = c("tomato1","royalblue1")) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill=group), alpha = .2) +
  geom_line() + theme_minimal()+
  labs(x = "Soil moisture (%)", y = "Total abundance")+ theme(plot.title = element_text(hjust = 0.5)) + 
  theme(legend.position="none",legend.text=element_text(size=15),legend.direction='horizontal',plot.title = element_text(hjust=0.5),text=element_text(family="Arial"))+
  theme(text=element_text(family="Arial")) + 
  theme(axis.title.y = element_text(margin=margin(r=60)), axis.title.x = element_text(margin=margin(t=20))) + 
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "transparent",colour = NA),
    plot.background = element_rect(fill = "transparent",colour = NA))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))
```

# Taxonomic Richness

We also used a generalized linear model with a Poisson distribution for Taxonomic Richness (the number of distinct taxonomic units, generally at Order level, found in each plot on each collection date).

taxa.mod <- glm(taxa ~ Avg..Soil.moisture * Plot.Type, family = poisson, data=soil.data)

Violin plot:

```{r, echo=FALSE, message = FALSE, warning = FALSE}
ggplot(soil.data, aes(x = Plot.Type, y = taxa, color = Plot.Type, group = Plot.Type)) + 
  scale_fill_manual(values = c("royalblue1", "tomato1")) + scale_color_manual(values = c("royalblue1", "tomato1")) +
  geom_violin(trim=FALSE)+  theme_minimal()+ labs(y= "Total abundance", x = "")+ geom_boxplot(size=0.1) +
  theme(plot.title = element_text(hjust = 0.5)) + stat_summary(fun=mean, geom="point") + stat_summary(fun.data = mean_se, geom = "errorbar", aes())+
  theme(legend.position="none",legend.text=element_text(),legend.direction='horizontal',plot.title = element_text(hjust=0.5),text=element_text(family="Arial"))+
  scale_x_discrete(labels= xaxis) +
  theme(text=element_text(family="Arial")) + 
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "transparent",colour = NA),
    plot.background = element_rect(fill = "transparent",colour = NA))+
  scale_y_continuous(labels=comma)+
    theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))

```

Effects plot:

```{r, echo=FALSE, message = FALSE, warning = FALSE}
taxa.effects <- ggpredict(taxa.mod, terms = c("Avg..Soil.moisture", "Plot.Type"))

ggplot(taxa.effects, aes(x = x, y = predicted, colour = group)) + 
  scale_fill_manual(values = c("tomato1","royalblue1")) + scale_color_manual(values = c("tomato1","royalblue1")) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill=group), alpha = .2) +
  geom_line() + theme_minimal()+
  labs(x = "Soil moisture (%)", y = "Total abundance")+ theme(plot.title = element_text(hjust = 0.5)) + 
  theme(legend.position="none",legend.text=element_text(size=15),legend.direction='horizontal',plot.title = element_text(hjust=0.5),text=element_text(family="Arial"))+
  theme(text=element_text(family="Arial")) + 
  theme(axis.title.y = element_text(margin=margin(r=60)), axis.title.x = element_text(margin=margin(t=20))) + 
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "transparent",colour = NA),
    plot.background = element_rect(fill = "transparent",colour = NA))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))

```

# Shannon Diversity

To find the right linear model we started by testing models that paired plot type with date, day from start, season, and soil moisture as fixed effects and Plot.pair as a random effect using the lmer() function from the package lme4. 

```{r, message=FALSE, results = 'hide'}
#date and plot type
model1 <- lmer(shannon ~ Date * Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)
model2 <- lmer(shannon ~ Date + (1|Plot.pair), REML = FALSE, data=soil.data,)
model3 <- lmer(shannon ~ Date + Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)
model4 <- lmer(shannon ~ Date:Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)
model5 <- lmer(shannon ~ Date + Plot.Type + Date:Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)      #should be same as model1

#day from start instead of date
model6 <- lmer(shannon ~ Day + (1|Plot.pair), REML = FALSE, data=soil.data,)
model7 <- lmer(shannon ~ Day + Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)
model8 <- lmer(shannon ~ Day:Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)
model9 <- lmer(shannon ~ Day*Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)


#season and plot type
model10 <- lmer(shannon ~ Season + (1|Plot.pair), REML = FALSE, data=soil.data,)
model11 <- lmer(shannon ~ Season + Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)
model12 <- lmer(shannon ~ Season:Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)
model13 <- lmer(shannon ~ Season * Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)

#moisture and plot type
model14 <- lmer(shannon ~ Avg..Soil.moisture + (1|Plot.pair), REML = FALSE, data=soil.data,)
model15 <- lmer(shannon ~ Avg..Soil.moisture + Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)
model16 <- lmer(shannon ~ Avg..Soil.moisture:Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)
model17 <- lmer(shannon ~ Avg..Soil.moisture * Plot.Type + (1|Plot.pair), REML = FALSE, data=soil.data,)

#to avoid overfitting, need a simpler model...


```

[I don't remember how we got from the various tests and experiments of models to the final one we used. I have a draft R script documenting the original tests (shown above), but these are all with the lmer() function, rather than lm() which we use in the final model.]

I tested and compared models using the Akaike information criterion (AIC), and ultimately chose the following structure: 

```{r, message=TRUE, results = 'show'}
shannon.mod <- lm(shannon ~ Avg..Soil.moisture * Plot.Type, data=soil.data,)
```
